---
title: "CDS 101 - Final Project: Structure-Based Screening for Drug-Induced Liver Injury"
subtitle: "Predicting drug-related liver injury concern from molecular structure (entire workflow in R)"
author: "Zamir Osmanzai, Usman Saleem (Usaleem16), Kebron (bizabebron-ops)"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    css: "gmu-cds101.css"
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 4
)
```

# Problem Definition

Drug-induced liver injury (DILI) means liver damage that appears to be caused or worsened by a medication. It is an important reason drugs fail in development or are withdrawn after approval. This project asks a narrower, technical question: can we guess a literature-based “elevated concern” vs. “lower concern” label from a drug’s chemical structure alone? A computer model that reads structure cannot replace animal studies, clinical trials, or expert review; at best it can flag molecules for closer human follow-up.

This report follows the usual CDS 101 arc: state the question, describe the data, clean and prepare it, explore patterns with plots, fit models with simple baselines, and say clearly what we can and cannot claim.

### Research questions (in plain terms)

1. Using only information we derive in R from each drug’s structure, how well can we predict the expert-style yes/no liver-injury concern label in the dataset?
2. Which kind of numeric summary of the structure works best? We compare three options: (a) a long fingerprint-a pattern of 0/1 bits that encodes small structural motifs around each atom; (b) a shorter list of whole-molecule numbers such as size, polarity, and flexibility; (c) both together. We mainly judge quality by ROC-AUC (how well the model ranks high-risk drugs above low-risk ones) and also look at precision and recall for calling “elevated concern.”

### Objectives

- Use a public table of drugs with SMILES strings (text line drawings of chemistry) and a binary concern label.
- Turn each SMILES into columns of numbers in R using the Chemistry Development Kit through the `rcdk` package (script `scripts/featurize_dili.R`).
- Fit several standard classifiers in R: a dumb baseline that always guesses the most common training label; logistic regression with ridge penalty (package `glmnet`, orchestrated by `tidymodels`) on three feature sets; and a random forest (`ranger`) on the combined features. We pick tuning settings with 5-fold cross-validation on the training portion only, then measure performance one time on a separate 25% test portion we set aside at the start.
- Be upfront about noise in labels, what the model does not see (dose, patients, metabolism), and why this is not a clinical decision tool.

### Assumptions

- The 0/1 labels summarize messy literature evidence; they are useful for learning but are not perfect ground truth about real-world liver risk.
- We read chemistry from SMILES as given; we do not explore every alternate salt form or molecular shape (a real safety pipeline would dig deeper).
- We randomly split drugs into train and test while keeping the proportion of “elevated” labels similar in each part. That matches typical classroom practice; a stricter “scaffold” split (keeping similar-looking drugs together) would stress-test new chemistry-we note that as a limitation later.

# Data Acquisition & Description

We use a gold-standard table shared by the open [DILI](https://github.com/srijitseal/DILI) project (Seal *et al.* and related expert curation). Each row is one drug: a SMILES string (a standard text encoding of its molecular structure) and a `TOXICITY` column that is 1 if the curators flagged elevated liver-injury concern and 0 for lower concern in their scheme. The raw spreadsheet lives in this repo at `data/DILI_Goldstandard_1111.csv`.

The Therapeutics Data Commons lists a related DILI task (on the order of 475 drugs) that many machine-learning papers use for benchmarks ([TDC toxicity tasks](https://tdcommons.ai/single_pred_tasks/tox/)). We mention it so readers can place our work in context; our numbers come from the larger CSV above because it is easy to download and rerun from a plain file.

```{r load-raw-meta}
library(readr)
library(dplyr)
library(forcats)

raw_dili <- read_csv("data/DILI_Goldstandard_1111.csv", show_col_types = FALSE)
glimpse(raw_dili)
```

```{r counts-raw}
raw_dili |>
  summarise(
    n_rows = n(),
    n_positive = sum(TOXICITY == 1, na.rm = TRUE),
    prevalence = mean(TOXICITY == 1, na.rm = TRUE)
  )
```

### Ethical and bias considerations

- The yes/no labels blend many kinds of evidence (case reports, labels, different patient groups). The drugs that appear in public databases are not a random sample of all chemistry, so performance here may not transfer to every population or drug class.
- This notebook is a classroom exercise in prediction from structure. It is not a medical device, not a regulatory submission, and not something to use to decide whether a real patient or program should take or avoid a drug without expert review.

# Data Cleaning & Preprocessing

### Step 1 - Turn structures into numbers (done once in R, outside this document)

To keep knitting quick, we pre-build a file of numeric features: `data/dili_modeling_features.csv.gz`. From the project folder, run `Rscript scripts/featurize_dili.R` (needs Java and the `rcdk` package; see `data/README.md`). That script:

- Reads the raw table (`smiles_r`, `TOXICITY`).
- Skips SMILES that do not parse into a valid molecule.
- Picks a standard (“canonical”) form of each SMILES and keeps one row per unique structure so duplicate drawings of the same drug do not inflate scores.
- Writes a compressed CSV of numbers plus the canonical SMILES and the integer label.

### Step 2 - What each column family means (read into this report)

- `desc_*` columns (whole-molecule summaries). Examples: molecular weight (how heavy the molecule is), XLogP (how fat-loving vs. water-loving it tends to be), TPSA (polar surface area-a rough sense of how “sticky” polar parts are), counts of hydrogen-bond donors/acceptors, rotatable bonds (flexibility), rings, aromatic atom count (stored under the name `desc_NumAromaticRings` for pipeline compatibility-see `data/README.md`), fraction sp³ carbons (a shape/complexity cue), heavy atoms, heteroatoms, and Bertz complexity (a size-and-branching complexity score). `desc_qed` is a slot for a “drug-likeness” score; the bundled R script leaves it blank (missing) because the CDK path we use does not compute QED. If your saved feature file shows real QED values, that file was likely built with a different featurization run-your plots will still run.
- `fp_*` columns (fingerprint bits). Each column is 0 or 1 for a tiny structural pattern. Together they form a circular fingerprint in the ECFP family (a common choice in cheminformatics; see Rogers & Hahn, 2010, in the References).

### Step 3 - Before fitting models (inside `tidymodels`)

We drop predictors that never change (useless for fitting), then subtract the mean and divide by the standard deviation for each numeric predictor so large and small columns are on a similar scale. The outcome column `dILI` is a category with plain-language levels `low_concern` and `elevated_concern`, copied from the raw 0/1 label.

```{r load-features}
feat_path <- "data/dili_modeling_features.csv.gz"
stopifnot(file.exists(feat_path))

dili <- read_csv(feat_path, show_col_types = FALSE) |>
  mutate(
    dILI = factor(
      label,
      levels = c(0, 1),
      labels = c("low_concern", "elevated_concern")
    )
  )

desc_cols <- names(dili)[startsWith(names(dili), "desc_")]
fp_cols <- names(dili)[startsWith(names(dili), "fp_")]

dili |>
  summarise(
    n = n(),
    n_positive = sum(dILI == "elevated_concern"),
    prevalence = mean(dILI == "elevated_concern")
  )
```

After cleaning, the modeling table has `r nrow(dili)` unique drugs, `r length(desc_cols)` whole-molecule descriptor columns, and `r length(fp_cols)` fingerprint bit columns (plus `smiles_canon` so you can trace each row back to a structure).
