---
title: "CDS 101 - Final Project: Structure-Based Screening for Drug-Induced Liver Injury"
subtitle: "Predicting drug-related liver injury concern from molecular structure (entire workflow in R)"
author: "Zamir Osmanzai, Usman Saleem (Usaleem16), Kebron (bizabebron-ops)"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    css: "gmu-cds101.css"
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 4
)
```

# Problem Definition

Drug-induced liver injury (DILI) means liver damage that appears to be caused or worsened by a medication. It is an important reason drugs fail in development or are withdrawn after approval. This project asks a narrower, technical question: can we guess a literature-based “elevated concern” vs. “lower concern” label from a drug’s chemical structure alone? A computer model that reads structure cannot replace animal studies, clinical trials, or expert review; at best it can flag molecules for closer human follow-up.

This report follows the usual CDS 101 arc: state the question, describe the data, clean and prepare it, explore patterns with plots, fit models with simple baselines, and say clearly what we can and cannot claim.

### Research questions (in plain terms)

1. Using only information we derive in R from each drug’s structure, how well can we predict the expert-style yes/no liver-injury concern label in the dataset?
2. Which kind of numeric summary of the structure works best? We compare three options: (a) a long fingerprint-a pattern of 0/1 bits that encodes small structural motifs around each atom; (b) a shorter list of whole-molecule numbers such as size, polarity, and flexibility; (c) both together. We mainly judge quality by ROC-AUC (how well the model ranks high-risk drugs above low-risk ones) and also look at precision and recall for calling “elevated concern.”

### Objectives

- Use a public table of drugs with SMILES strings (text line drawings of chemistry) and a binary concern label.
- Turn each SMILES into columns of numbers in R using the Chemistry Development Kit through the `rcdk` package (script `scripts/featurize_dili.R`).
- Fit several standard classifiers in R: a dumb baseline that always guesses the most common training label; logistic regression with ridge penalty (package `glmnet`, orchestrated by `tidymodels`) on three feature sets; and a random forest (`ranger`) on the combined features. We pick tuning settings with 5-fold cross-validation on the training portion only, then measure performance one time on a separate 25% test portion we set aside at the start.
- Be upfront about noise in labels, what the model does not see (dose, patients, metabolism), and why this is not a clinical decision tool.

### Assumptions

- The 0/1 labels summarize messy literature evidence; they are useful for learning but are not perfect ground truth about real-world liver risk.
- We read chemistry from SMILES as given; we do not explore every alternate salt form or molecular shape (a real safety pipeline would dig deeper).
- We randomly split drugs into train and test while keeping the proportion of “elevated” labels similar in each part. That matches typical classroom practice; a stricter “scaffold” split (keeping similar-looking drugs together) would stress-test new chemistry-we note that as a limitation later.

# Data Acquisition & Description

We use a gold-standard table shared by the open [DILI](https://github.com/srijitseal/DILI) project (Seal *et al.* and related expert curation). Each row is one drug: a SMILES string (a standard text encoding of its molecular structure) and a `TOXICITY` column that is 1 if the curators flagged elevated liver-injury concern and 0 for lower concern in their scheme. The raw spreadsheet lives in this repo at `data/DILI_Goldstandard_1111.csv`.

The Therapeutics Data Commons lists a related DILI task (on the order of 475 drugs) that many machine-learning papers use for benchmarks ([TDC toxicity tasks](https://tdcommons.ai/single_pred_tasks/tox/)). We mention it so readers can place our work in context; our numbers come from the larger CSV above because it is easy to download and rerun from a plain file.

```{r load-raw-meta}
library(readr)
library(dplyr)
library(forcats)

raw_dili <- read_csv("data/DILI_Goldstandard_1111.csv", show_col_types = FALSE)
glimpse(raw_dili)
```

```{r counts-raw}
raw_dili |>
  summarise(
    n_rows = n(),
    n_positive = sum(TOXICITY == 1, na.rm = TRUE),
    prevalence = mean(TOXICITY == 1, na.rm = TRUE)
  )
```

### Ethical and bias considerations

- The yes/no labels blend many kinds of evidence (case reports, labels, different patient groups). The drugs that appear in public databases are not a random sample of all chemistry, so performance here may not transfer to every population or drug class.
- This notebook is a classroom exercise in prediction from structure. It is not a medical device, not a regulatory submission, and not something to use to decide whether a real patient or program should take or avoid a drug without expert review.

# Data Cleaning & Preprocessing

### Step 1 - Turn structures into numbers (done once in R, outside this document)

To keep knitting quick, we pre-build a file of numeric features: `data/dili_modeling_features.csv.gz`. From the project folder, run `Rscript scripts/featurize_dili.R` (needs Java and the `rcdk` package; see `data/README.md`). That script:

- Reads the raw table (`smiles_r`, `TOXICITY`).
- Skips SMILES that do not parse into a valid molecule.
- Picks a standard (“canonical”) form of each SMILES and keeps one row per unique structure so duplicate drawings of the same drug do not inflate scores.
- Writes a compressed CSV of numbers plus the canonical SMILES and the integer label.

### Step 2 - What each column family means (read into this report)

- `desc_*` columns (whole-molecule summaries). Examples: molecular weight (how heavy the molecule is), XLogP (how fat-loving vs. water-loving it tends to be), TPSA (polar surface area-a rough sense of how “sticky” polar parts are), counts of hydrogen-bond donors/acceptors, rotatable bonds (flexibility), rings, aromatic atom count (stored under the name `desc_NumAromaticRings` for pipeline compatibility-see `data/README.md`), fraction sp³ carbons (a shape/complexity cue), heavy atoms, heteroatoms, and Bertz complexity (a size-and-branching complexity score). `desc_qed` is a slot for a “drug-likeness” score; the bundled R script leaves it blank (missing) because the CDK path we use does not compute QED. If your saved feature file shows real QED values, that file was likely built with a different featurization run-your plots will still run.
- `fp_*` columns (fingerprint bits). Each column is 0 or 1 for a tiny structural pattern. Together they form a circular fingerprint in the ECFP family (a common choice in cheminformatics; see Rogers & Hahn, 2010, in the References).

### Step 3 - Before fitting models (inside `tidymodels`)

We drop predictors that never change (useless for fitting), then subtract the mean and divide by the standard deviation for each numeric predictor so large and small columns are on a similar scale. The outcome column `dILI` is a category with plain-language levels `low_concern` and `elevated_concern`, copied from the raw 0/1 label.

```{r load-features}
feat_path <- "data/dili_modeling_features.csv.gz"
stopifnot(file.exists(feat_path))

dili <- read_csv(feat_path, show_col_types = FALSE) |>
  mutate(
    dILI = factor(
      label,
      levels = c(0, 1),
      labels = c("low_concern", "elevated_concern")
    )
  )

desc_cols <- names(dili)[startsWith(names(dili), "desc_")]
fp_cols <- names(dili)[startsWith(names(dili), "fp_")]

dili |>
  summarise(
    n = n(),
    n_positive = sum(dILI == "elevated_concern"),
    prevalence = mean(dILI == "elevated_concern")
  )
```

After cleaning, the modeling table has `r nrow(dili)` unique drugs, `r length(desc_cols)` whole-molecule descriptor columns, and `r length(fp_cols)` fingerprint bit columns (plus `smiles_canon` so you can trace each row back to a structure).

# Exploratory Data Analysis (EDA)

## Class balance

```{r eda-balance, fig.cap="How many drugs fall in each label group after featurization (expert-style elevated vs. lower concern)."}
library(ggplot2)

dili |>
  count(dILI) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(dILI, n, fill = dILI)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = scales::percent(pct, accuracy = 0.1)), vjust = -0.35, size = 4) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    x = NULL,
    y = "Count",
    title = "Outcome distribution (elevated concern is the majority class here)"
  ) +
  theme_minimal(base_size = 13)
```

In this table, “elevated concern” is usually the majority class (often well above half of the rows). That means a model could get a decent accuracy by almost always guessing “elevated,” even if it is not truly informative. So we lean on metrics that are harder to “game” that way-especially ROC-AUC (ranking quality) and precision-recall, plus sensitivity (fraction of elevated cases caught) and specificity (fraction of lower-concern cases correctly left alone) on the held-out test drugs the model never trained on.

## Descriptor distributions by class

```{r eda-box, fig.cap="Example numeric summaries of chemistry, split by label. Overlap is expected when a single number tries to stand in for a complex safety outcome."}
toplot <- intersect(
  c("desc_MolWt", "desc_MolLogP", "desc_TPSA", "desc_qed"),
  desc_cols
)
# Omit columns that are entirely missing (e.g. QED when not computed in R featurizer)
toplot <- toplot[vapply(dili[toplot], function(x) any(!is.na(x)), logical(1))]

desc_labs <- c(
  desc_MolWt = "Molecular weight",
  desc_MolLogP = "XLogP",
  desc_TPSA = "TPSA",
  desc_qed = "QED (drug-likeness)"
)

dili |>
  select(dILI, all_of(toplot)) |>
  tidyr::pivot_longer(-dILI, names_to = "descriptor", values_to = "value") |>
  mutate(
    descriptor = forcats::fct_inorder(factor(descriptor)),
    descriptor_lab = unname(desc_labs[as.character(descriptor)])
  ) |>
  mutate(descriptor_lab = forcats::fct_inorder(factor(descriptor_lab))) |>
  ggplot(aes(dILI, value, fill = dILI)) +
  geom_boxplot(outlier.alpha = 0.35, show.legend = FALSE) +
  facet_wrap(~descriptor_lab, scales = "free_y") +
  scale_fill_brewer(palette = "Pastel2") +
  labs(
    title = "Selected molecular descriptors vs. label",
    x = NULL,
    y = "Value"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

How to read the boxplots: for each drug property on its own, the two label groups overlap a lot. That is normal: liver injury depends on biology and dosing we do not feed into the model, and the labels themselves are imperfect. The hope is that many weak clues together (through the models below) do better than any single plot.

## Linear correlations among descriptors

```{r eda-corr, fig.cap="Correlation heatmap: brighter colors mean two descriptors tend to increase or decrease together across drugs."}
library(corrplot)

cor_mat <- dili |>
  select(all_of(desc_cols)) |>
  cor(use = "pairwise.complete.obs")

corrplot(
  cor_mat,
  method = "color",
  type = "upper",
  tl.cex = 0.7,
  mar = c(0, 0, 2, 0),
  title = "Which descriptors move together? (many pairs are correlated)"
)
```

Many descriptors rise and fall together (bright off-diagonal patches). When predictors are correlated, a giant vanilla logistic regression can overreact to noise. Ridge logistic regression (what we use) is a standard fix: it shrinks coefficients toward zero so the model stays stable when there are many related columns-especially important for the long fingerprint block.

# Visualization Quality and Storytelling

- Side-by-side boxplots compare each numeric property for “lower” vs. “elevated” concern.  
- The heatmap shows which properties move together.  
- ROC curves show, across all possible cutoff rules, the tradeoff between true alarms (vertical axis) and false alarms (horizontal axis) when ranking drugs by predicted risk. A curve closer to the upper-left is better; the dashed diagonal is random guessing.  
- Precision-recall curves focus on the elevated class: if we flag the model’s top guesses, how many are correct (precision) and what fraction of all elevated drugs did we catch (recall)?  
- The horizontal bar chart from the random forest highlights which inputs the ensemble leaned on most (treat as hints, not causal proof).  

We use ColorBrewer colors and plain-language facet titles where we can so figures read quickly on screen or on paper.

# Modeling Approach

Each drug is either `elevated_concern` or `low_concern`. Models output a probability of elevated concern and a predicted class once we pick a cutoff (the software defaults handle that for the confusion matrix).

### Baselines and comparators

1. Majority baseline. Always predict whichever label appeared most often in the training set, and assign every drug the same baseline probability equal to the training fraction of elevated cases. This is a sanity check: a serious model should beat “always guess the popular answer.”
2. Ridge logistic regression (R packages `glmnet` and `tidymodels`). Think “logistic regression, but the math penalizes huge coefficients so the fit does not chase noise.” We try 12 penalty strengths on a log scale, using 5-fold cross-validation on the training 75% only to pick a strength, then refit on the full training 75% and score the held-out 25%.
3. Random forest (`ranger`). Many decision trees vote; each tree sees a random subset of rows and columns, which often helps with nonlinear patterns. We tune two knobs (`mtry`, `min_n`) on the same 5-fold training-only cross-validation, then evaluate on the same held-out test split.

### Three ways we feed the numbers in

- Descriptors only - short list, easier to relate to chemistry intuition.  
- Fingerprints only - long bit vector, good at capturing local shape in the graph.  
- Both together - asks whether the two views add information.

### How the data are split

We use `rsample::initial_split` to put 75% of drugs in training and 25% in testing, keeping roughly the same mix of labels in each part (stratified split). Parameter choices are tuned only on training; the test drugs are touched once at the end for the tables and plots. `set.seed(2026)` makes the split and forest randomness repeatable on a fresh run.

# Model Implementation & Evaluation

The code chunks below run the steps just described, then print a metric table, ROC and precision-recall plots, variable importance for the forest, and a confusion matrix for whichever model had the best test ROC-AUC (excluding the trivial majority baseline).

```{r modeling-prep}
library(tidymodels)
library(workflows)
library(recipes)
library(parsnip)
library(tune)
library(dials)
library(yardstick)
library(purrr)
library(ranger)
library(glmnet)

set.seed(2026)

data_model <- dili |>
  select(dILI, all_of(desc_cols), all_of(fp_cols))

split_obj <- initial_split(data_model, prop = 0.75, strata = dILI)
train_data <- training(split_obj)
test_data <- testing(split_obj)

make_recipe <- function(train_df) {
  recipe(dILI ~ ., data = train_df) |>
    step_zv(all_predictors()) |>
    step_normalize(all_numeric_predictors())
}

lr_spec <- logistic_reg(penalty = tune(), mixture = 0) |>
  set_engine("glmnet") |>
  set_mode("classification")

penalty_grid <- grid_regular(
  penalty(range = c(-5, 0)),
  levels = 12
)

tune_lr <- function(feature_cols, label) {
  train_slice <- train_data |> select(dILI, all_of(feature_cols))
  folds <- vfold_cv(train_slice, v = 5, strata = dILI)
  rec <- make_recipe(train_slice)
  wf <- workflow() |>
    add_recipe(rec) |>
    add_model(lr_spec)

  tuned <- tune_grid(
    wf,
    resamples = folds,
    grid = penalty_grid,
    metrics = metric_set(roc_auc, pr_auc, accuracy),
    control = control_grid(save_pred = FALSE, verbose = FALSE)
  )

  best_params <- select_best(tuned, metric = "roc_auc")
  final_wf <- finalize_workflow(wf, best_params)
  fit_train <- fit(final_wf, data = train_data |> select(dILI, all_of(feature_cols)))

  preds <- predict(fit_train, test_data |> select(dILI, all_of(feature_cols)), type = "prob")
  hard <- predict(fit_train, test_data |> select(dILI, all_of(feature_cols)), type = "class")

  tibble::tibble(
    truth = test_data$dILI,
    .pred_low_concern = preds$.pred_low_concern,
    .pred_elevated_concern = preds$.pred_elevated_concern,
    .pred_class = hard$.pred_class,
    model = label
  )
}

roc_rows <- function(pred_tbl) {
  m <- pred_tbl$model[1]
  roc_curve(pred_tbl, truth, .pred_elevated_concern, event_level = "second") |>
    mutate(model = m)
}

pr_rows <- function(pred_tbl) {
  m <- pred_tbl$model[1]
  pr_curve(pred_tbl, truth, .pred_elevated_concern, event_level = "second") |>
    mutate(model = m)
}
```

```{r modeling-tune-glmnet, cache=TRUE}
pred_desc <- tune_lr(desc_cols, "Penalized LR (descriptors)")
pred_fp <- tune_lr(fp_cols, "Penalized LR (fingerprints)")
pred_both <- tune_lr(c(desc_cols, fp_cols), "Penalized LR (combined)")
```

```{r modeling-rf, cache=TRUE}
rf_rec <- recipe(dILI ~ ., data = train_data) |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors())

rf_spec <- rand_forest(
  mtry = tune(),
  min_n = tune(),
  trees = 400
) |>
  set_engine("ranger", importance = "permutation", seed = 2026) |>
  set_mode("classification")

rf_wf <- workflow() |>
  add_recipe(rf_rec) |>
  add_model(rf_spec)

p_rf <- ncol(train_data) - 1L
mtry_hi <- max(2L, min(40L, 3L * floor(sqrt(p_rf))))
rf_grid <- grid_regular(
  mtry(range = c(2, mtry_hi)),
  min_n(range = c(2, 20)),
  levels = 3
)

cv_rf <- vfold_cv(train_data, v = 5, strata = dILI)

rf_tune <- tune_grid(
  rf_wf,
  resamples = cv_rf,
  grid = rf_grid,
  metrics = metric_set(roc_auc, pr_auc, accuracy),
  control = control_grid(save_pred = FALSE, verbose = FALSE)
)

rf_best <- finalize_workflow(rf_wf, select_best(rf_tune, metric = "roc_auc"))
rf_fit <- fit(rf_best, data = train_data)

rf_prob <- predict(rf_fit, test_data, type = "prob")
rf_cls <- predict(rf_fit, test_data, type = "class")

pred_rf <- test_data |>
  dplyr::select(truth = dILI) |>
  dplyr::bind_cols(rf_prob, rf_cls) |>
  dplyr::mutate(model = "Random forest (combined)")
```
